Machine learning guided batched design\\ of a bacterial Ribosome Binding Site

### 2021-11-21 paper/main.tex
mengyan edits on r2 and ee colors

\title{Multi-armed bandit \mengyan{(Bayesian?)} optimisation\\
of bacterial Ribosome Binding Site}

@@

Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group shows the highest TIR for that quantile.
The gradual increase for all quantiles can be observed for all Bandit groups, suggesting that the algorithms have a better understanding of the experimental space given more data.
The decreased result in the 0.9th quantile compared to the maximum value for the Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for that round compared to others.
We see this effect in Figure \ref{fig: Swarmplot and Quantplot}C (with details shown in Figure  \ref{fig:exploitationvsexplorationdetail}), where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
Those with a high predicted mean are coloured blue and represent exploitation, those coloured red are with high predicted uncertainty and represent exploration.
\hl{Since we expect that an RBS sequence chosen at random will have low TIR
(as shown by UNI and PPM), this implies that most of the exploration will result in low TIR.
Our results confirm that
RBSs with high TIRs tend to come from exploitation of the design space,
whereas the exploration points give relatively low TIRs. Note that the exploration is necessary to expand our knowledge of unknown parts of the design space.}\\

@@

\subsection{LEARN: Prediction of RBS performance}
\label{sec:gp-results}

@@

The regression task in the DBTL cycle is more challenging than the large-scale data based regression tasks \cite{}.
At the early iterations, we have a limited number of data points and relatively high variants due to the measurement noise. 
Fortunately, since the predictions are only used for recommendations in the Design part, instead of precise predictions, we only need to provide a valid ranking for both predicted TIR and uncertainties, and thus the ranking of UCB scores. 
So we provide the evaluation of Learn part in terms of a ranking-based Spearman correlation coefficient, which has been shown as a more suitable evaluation metric when the prediction is used for recommendation tasks than $R^2$ or Pearman correlation coefficient
\cite{Schober2018,  Kang2019}.
\mengyan{Do we even want to put R2 score in?}\\

Figure \ref{fig: Scatterplot} shows how our implementation of the Gaussian Process algorithm performed in terms of predictions in each round.  
In (A)(B)(C), we show at round $t = 1,2,3$ respectively, we train our predictor based on the previously obtained data (round 0 to $t-1$) and show the predictions on both the training data (orange) and recommendations suggested by the Design part (i.e. test data, blue). 
We reported the Spearman and $R^2$ score in each plot's title on test data only. 
X-axis and y-axis are true TIR from our Test part of the DBTL cycle and the predicted TIR from our predictor.
In (D), we train our predictor on randomly chosen 80\% of all available data, and test on the rest 20\% data. 
As expected, the predictions in Round 0 were poor due to the use of approximate data. 
The predictions improved for the subsequent rounds, 
% from R\textsuperscript{2} (coefficient of determination) of 0.067 for Round 0 to R\textsuperscript{2} of 0.213 for Round 3.
the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.
\mengyan{We also show $R_2$ score for the reader's interest, which quantifies value-based prediction evaluation. Similarly, the $R_2$ increases along with rounds increase. }
% Figure \ref{fig:scatter abc1 TT.} shows the performance of predictions for individual rounds.
\\

@@

\begin{figure}[!ht]
    \centering
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_0.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_1.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_2.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_3.pdf}
    \end{subfigure}
    \caption{\textbf{Performance of the prediction algorithm (no kernel normalisation)}. The scatter plots A-D show the performance of our prediction algorithm calculated after each round.
    Note that the TIR values are normalised according to the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in Figure \ref{fig: Swarmplot and Quantplot}.
    The $R^2$ value and Spearman correlation coefficient (with corresponding p-value) are provided for each plot.
    The p-value here is for the null hypothesis stating that two sets of data are uncorrelated.
    }
    \label{fig: Scatterplot}
\end{figure}

@@

Our approach has shown that machine learning is able to correctly detect and exploit rules of biological design that otherwise require substantial time and experiments to uncover.
For example, our algorithm has correctly identified the correlation of high guanine content in the RBS with high TIR.
We have achieved this despite the relatively low R\textsuperscript{2} values for our predictions.
Notably, other machine learning methods used in synthetic biology on small data sets were also shown to produce useful recommendations despite low R\textsuperscript{2} values \cite{Radivojevic2020, Opgenorth2019}.
Additionally, it has been shown that when the prediction is used for recommendation tasks like the one considered in this paper, ranking-based metric like Spearman correlation coefficient is a better evaluation metric
\cite{Schober2018,  Kang2019}.
Finally, our predictor Gaussian Process regression model, compared with previously described calculators using a deterministic thermodynamic approach, is able to show the uncertainty of the predictions, which can be used by our Bandit algorithm to give better recommendations.\\

>>>>

\title{Machine learning guided batched design\\
of a bacterial Ribosome Binding Site}

@@

Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group shows the highest TIR for that quantile.
The gradual increase for all quantiles can be observed for all Bandit groups, suggesting that the algorithms have a better understanding of the experimental space given more data.
The decreased result in the 0.9th quantile compared to the maximum value for the Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for that round compared to others.
We see this effect in Figure \ref{fig: Swarmplot and Quantplot}C (with details shown in Supplementary \ref{subsec: Exploration-exploitation visualisation} and Figure  \ref{fig:exploitationvsexplorationdetail}), where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
Those with a high predicted mean are coloured blue and represent exploitation, those coloured red are with high predicted uncertainty and represent exploration.
\hl{Since we expect that an RBS sequence chosen at random will have low TIR
(as shown by UNI and PPM), this implies that most of the exploration will result in low TIR.
Our results confirm that
RBSs with high TIRs tend to come from exploitation of the design space,
whereas the exploration points give relatively low TIRs. Note that the exploration is necessary to expand our knowledge to unknown parts of the design space.}\\

@@

\subsection{LEARN: Prediction of RBS performance}
\label{sec:gp-results}
\label{subsec: results LEARN}

@@

The regression task in the DBTL cycle is more challenging than the large-scale data based regression tasks.
At the early iterations, we have a limited number of data points and relatively high variants due to the measurement noise. 
Fortunately, since the predictions are only used for recommendations in the Design part, instead of precise predictions, we only need to provide a valid ranking for both predicted TIR and uncertainties, and thus the ranking of UCB scores. 
So we provide the evaluation of Learn part in terms of a ranking-based Spearman correlation coefficient, which has been shown as a more suitable evaluation metric when the prediction is used for recommendation tasks than $R^2$ or Pearman correlation coefficient
\cite{Schober2018,  Kang2019}.\\

Figure \ref{fig: Scatterplot} shows how our implementation of the Gaussian Process algorithm performed in terms of predictions in each round.  
In (A)(B)(C), we show at round $t = 1,2,3$ respectively, we train our predictor based on the previously obtained data (round 0 to $t-1$) and show the predictions on both the training data (orange) and recommendations suggested by the Design part (i.e. test data, blue). 
We reported the Spearman score in each plot's title on test data only. 
X-axis and y-axis are true TIR from our Test part of the DBTL cycle and the predicted TIR from our predictor.
In (D), we train our predictor on randomly chosen 80\% of all available data, and test on the rest 20\% data. 
As expected, the predictions in Round 0 were poor due to the use of approximate data. 
The predictions improved for the subsequent rounds, 
% from R\textsuperscript{2} (coefficient of determination) of 0.067 for Round 0 to R\textsuperscript{2} of 0.213 for Round 3.
the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.
% Figure \ref{fig:scatter abc1 TT.} shows the performance of predictions for individual rounds.
\\

@@

\begin{figure}[!ht]
    \centering
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_0.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_1.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_2.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_3.pdf}
    \end{subfigure}
    \caption{\textbf{Performance of the prediction algorithm (no kernel normalisation)}. The scatter plots A-D show the performance of our prediction algorithm calculated after each round.
    Note that the TIR values are normalised according to the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in Figure \ref{fig: Swarmplot and Quantplot}.
    The Spearman correlation coefficient (with corresponding p-value) are provided for each plot.
    The p-value here is for the null hypothesis stating that two sets of data are uncorrelated.
    }
    \label{fig: Scatterplot}
\end{figure}

@@

Our approach has shown that machine learning is able to correctly detect and exploit rules of biological design that otherwise require substantial time and experiments to uncover.
For example, our algorithm has correctly identified the correlation of high guanine content in the RBS with high TIR.
We have achieved this despite the relatively low Spearman scores for our predictions.
Notably, other machine learning methods used in synthetic biology on small data sets were also shown to produce useful recommendations despite low prediction scores \cite{Radivojevic2020, Opgenorth2019}.
% Additionally, it has been shown that when the prediction is used for recommendation tasks like the one considered in this paper, ranking-based metric like Spearman correlation coefficient is a better evaluation metric
% \cite{Schober2018,  Kang2019}.
Finally, our predictor Gaussian Process regression model, compared with previously described calculators using a deterministic thermodynamic approach, is able to show the uncertainty of the predictions, which can be used by our Bandit algorithm to give better recommendations.\\

### 2021-11-21 paper/supplementary.tex
mengyan edits on r2 and ee colors

@@

\begin{figure}[!ht]
   \centering
    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_0.pdf}
    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_1.pdf}
    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_2.pdf}
   \caption{\textbf{Exploration-Exploitation scores illustration.} We show the predictions for each sequence from Round 1-3 in orange (with the same settings as Figure \mbox{\ref{fig: Scatterplot}}).
   For a given round all the tested sequences are then projected along the diagonal line crossing (0,1) and (1,0) according to their predicted STD and mean TIR (the higher the predicted mean the closer the point is to the (1,0) coordinate and the higher the predicted STD the closer the point is to the (0,1) coordinate).
   Each point on the diagonal then, corresponds to an orthogonal projection (known as \emph{vector projection}) of an orange point.
   We do this to transform our predictions into a common basis (where the basis is the diagonal) and so by calculating where a given point falls on the diagonal we can show if its mean TIR or STD is higher.
   Finally, we colour those projected points using the RdBu spectrum where the points with high STD (exploration) are shown in red and those with high mean TIR (exploitation) are shown in blue, and these correspond to the points shown in Figure \mbox{\ref{fig: Swarmplot and Quantplot}C}.
   }
   \label{fig:exploitationvsexplorationdetail}
\end{figure}

\begin{figure}[!ht]
    \centering
    \includegraphics[scale = 0.35]{plots/Supplementary/scatter_bc1_TTF_0.pdf}
    \includegraphics[scale = 0.35]{plots/Supplementary/scatter_abc1_TT_1.pdf}
    \includegraphics[scale = 0.35]{plots/Supplementary/scatter_abc1_TT_2.pdf}
    % \includegraphics[scale = 0.4]{plots/Supplementary/scatter_abc1_TT_3.pdf}
    \caption{\textbf{Predictions in design pipeline in Round 1-3.} For Round 1, the kernel matrix is normalised based on known RBS sequences in Round 0.
    For Round 2-3, the kernel matrix is normalised based on all RBS sequences in the design space.
    Compared to the results shown in Figure \mbox{\ref{fig: Scatterplot}}, here we have applied kernel normalisation described in Section \mbox{\ref{supp: Normalisation of Kernel}}. Although the $R^2$ scores decrease considerably compared with Figure \mbox{\ref{fig: Scatterplot}}, the Spearman correlation remains similar.
    }
    \label{fig:scatter abc1 TT.}
\end{figure}

>>>>

\subsection{Addition Results}

As stated in Section \ref{subsec: results LEARN}, we merely aimed to provide a valid ranking in Learn part due to the small amount of noisy data, which can be captured by the Spearman correlation coefficient. 
We show $R^2$ score for the reader's interest, which quantifies value-based prediction evaluation. The $R^2$ increases along with rounds increase.

\begin{table}[h!]
\caption{$R^2$ score for predictions in Figure \ref{fig: Scatterplot}.}
\centering
\begin{tabular}{|c|c|c|c|c|}
\hline
                  & \textbf{\begin{tabular}[c]{@{}c@{}}Train 0\\ Test 1\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Train 0-1\\ Test 2\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Train 0-2\\ Test 3\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Train 80\%\\ Test 20\%\end{tabular}} \\ \hline
\textbf{$R^2$}       & 0.067                                                             & 0.226                                                               & 0.213                                                               & 0.269                                                                   \\ \hline
\textbf{Spearman} & 0.269                                                             & 0.468                                                               & 0.546                                                               & 0.479                                                                   \\ \hline
\end{tabular}
\end{table}

\subsection{Exploration-exploitation visualisation}
\label{subsec: Exploration-exploitation visualisation}

In this section, we show our method of exploration-exploitation visualisation displayed in Figure \ref{fig: Swarmplot and Quantplot}C.
Recall the UCB score of a sequence $\mathbf{x}_i$ at Round $t$ is defined as 
\begin{align*}
% \label{Eq: UCB}
    \mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i),
\end{align*}
where $\beta_t$ is a hyperparameter balancing the exploration and exploitation, 
$\mu_{t-1}(\mathbf{x}_i), \sigma_{t-1}(\mathbf{x}_i)$ are the predicted TIR and standard deviation (STD) at Round $t-1$ for the sequence $\mathbf{x}_i$.
The sequences with high predicted TIR address exploitation, while the sequences with high predicted STD address exploration. 
We show the predicted TIR (x-axis) v.s. predicted STD (y-axis) in Figure \ref{fig:exploitationvsexplorationdetail} in orange points.

To compare the predicted TIR and STD in the same dimension (basis), we project the (predicted TIR, predicted STD) pair to the diagonal line (the grey dashline) crossing (0,1) and (1,0).
The slope of the diagonal line is selected to weight the predicted TIR and STD equally. 
Each point on the diagonal then, corresponds to an orthogonal projection (known as \emph{vector projection}) of an orange point.
As a result, the higher the predicted TIR the closer the point is to the (1,0) coordinate, and the higher the predicted STD the closer the point is to the (0,1) coordinate.
Finally, we colour those projected points using the RdBu spectrum (according to either of the two dimension of the projected points), where the points with high STD (exploration) are shown in red and those with high mean TIR (exploitation) are shown in blue.
These colors are correspond to ones shown in Figure \mbox{\ref{fig: Swarmplot and Quantplot}C}.

% \mengyan{mengyan to re-write it}
% For a given round all the tested sequences are then projected along the diagonal line crossing (0,1) and (1,0) according to their predicted STD and mean TIR (the higher the predicted mean the closer the point is to the (1,0) coordinate and the higher the predicted STD the closer the point is to the (0,1) coordinate).
% Each point on the diagonal then, corresponds to an orthogonal projection (known as \emph{vector projection}) of an orange point.
% We do this to transform our predictions into a common basis (where the basis is the diagonal) and so by calculating where a given point falls on the diagonal we can show if its mean TIR or STD is higher.
% Finally, we colour those projected points using the RdBu spectrum where the points with high STD (exploration) are shown in red and those with high mean TIR (exploitation) are shown in blue.

\begin{figure}[!ht]
   \centering
    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_0.pdf}
    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_1.pdf}
    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_2.pdf}
   \caption{\textbf{Exploration-Exploitation scores illustration.} We show the predictions for each sequence from Round 1-3 in orange (with the same settings as Figure \mbox{\ref{fig: Scatterplot}}).
    We explain how the colors are generated in Section \ref{subsec: Exploration-exploitation visualisation}, which are correspond to the colors shown in Figure \mbox{\ref{fig: Swarmplot and Quantplot}C.}
   }
   \label{fig:exploitationvsexplorationdetail}
\end{figure}

@@

\begin{figure}[!ht]
    \centering
    \includegraphics[scale = 0.35]{plots/Supplementary/scatter_bc1_TTF_0.pdf}
    \includegraphics[scale = 0.35]{plots/Supplementary/scatter_abc1_TT_1.pdf}
    \includegraphics[scale = 0.35]{plots/Supplementary/scatter_abc1_TT_2.pdf}
    % \includegraphics[scale = 0.4]{plots/Supplementary/scatter_abc1_TT_3.pdf}
    \caption{\textbf{Predictions in design pipeline in Round 1-3.} For Round 1, the kernel matrix is normalised based on known RBS sequences in Round 0.
    For Round 2-3, the kernel matrix is normalised based on all RBS sequences in the design space.
    Compared to the results shown in Figure \mbox{\ref{fig: Scatterplot}}, here we have applied kernel normalisation described in Section \mbox{\ref{supp: Normalisation of Kernel}}. Although the precise value prediction looks different from Figure \mbox{\ref{fig: Scatterplot}}, the rankings provided in two figures are similar, which can be reflected by the similar Spearman correlation coefficients.
    }
    \label{fig:scatter abc1 TT.}
\end{figure}
